---
title: "Resolving names offline"
author: "Brian Maitner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resolving names offline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Status: beta

The offline engine is a new port of the published TNRS algorithm. Measured
against the web service over the 100 name benchmark in `tnrs_testfile`:

| | agree |
|---|---|
| `Name_matched` | 99 of 100 (97 exactly, two differing only in the hybrid marker) |
| `Accepted_name` | 87 of 100 |

The thirteen accepted-name differences are worth understanding, because most of
them are not errors.

**Five reflect newer data.** The service's backbone was built on 2024-01-17;
the local copy is WFO 2025-12 and WCVP v15. Where a treatment changed in
between, the local answer is the current one. `Warnstorfia trichophylla`
resolves to `Sarmentypnum trichophyllum` through the service, but WFO now
accepts `Warnstorfia trichophylla` and treats `Sarmentypnum trichophyllum` as
the synonym. `Chamaecrista conferta var. gurgueiana` and
`Lycium californicum subsp. carinatum` are both returned as accepted names by
the service and are synonyms in both current sources. For
`Engelhardia spicata var. colebrookeana` the service gives no accepted name at
all, where the local version gives one.

**Four are the two sources disagreeing.** WFO and WCVP do not always accept the
same name, and the two implementations follow different ones. `Bartlettina
macrocephala` is accepted in WFO and a synonym in WCVP; the service reports
WCVP's answer and the local version WFO's. Neither is wrong. Ask for both
sources and `Source_conflict` marks these.

**The rest are rank conventions**, such as `Gentianopsis barbata subsp. barbata`
against `var. barbata`, which is WFO and WCVP writing the same taxon at
different ranks.

**One difference is not in the local version's favour.** `Brachyphyllum
menendezii`, a fossil conifer genus in neither backbone, is matched by the
service to its family, `Araucariaceae`. The local version fuzzy-matches it to
`Trachyphyllum`, a moss genus. That is the false-positive pattern described
under "Names that are not plants" below, and here the service is the more
cautious of the two.

Beyond this benchmark, the engine has not been tested broadly across taxonomic
groups, across all the sources it can now read, or against the range of messy
input real datasets contain, and it has not been independently reviewed. The
animal sources in particular are very new. Treat the results as provisional:
for work where the answer matters, resolve a sample through `TNRS()` as well and
compare, and please report anything that disagrees.

## Why resolve names offline

`TNRS()` sends your names to the Taxonomic Name Resolution Service and returns
the server's answer. That is the right choice most of the time, but it needs an
internet connection, it needs the service to be up, and very large jobs are sent
in batches of 5000 with a round trip for each.

`TNRS_local()` runs the same matching algorithm on your own machine, against a
copy of the taxonomic sources that you download once. It works with no network,
it does not depend on the service being available, and there are no round trips.

The two are separate functions rather than one function with a switch. That is
deliberate: they do not always give the same answer, and it should always be
obvious from your code which one produced a result.

## One-off setup

```{r}
library(TNRS)

TNRS_local_build()
```

This downloads the World Flora Online backbone from its publisher and prepares
it for matching: about 116 MB, and a few minutes, leaving about 120 MB on disk.
You only need to do it again when you want a newer version of the source.

The downloaded archive is deleted once the data has been prepared, because
matching does not need it. If you expect to rebuild, keep it:

```{r}
TNRS_local_build(keep_archive = TRUE)
```

That roughly doubles the space used, and saves the download next time. Either
way the archive's size and checksum stay in the record, so removing it costs
nothing you would want to cite.

`TNRS_local()` uses that one source by default. If you want the second source as
well, so that you can compare the two the way the web service does, build it too:

```{r}
TNRS_local_build("wcvp")
```

Asking `TNRS_local()` for a source you have not built does not quietly start a
download. It tells you what is missing and the call that would fix it:

```
No local copy of: wcvp. (wfo is already built.)
Run TNRS_local_build("wcvp") once to download and prepare it (about 85 MB to download).
Or call this function again with build_missing = TRUE to do it now.
```

In an interactive session you can pass `build_missing = TRUE` and answer the
prompt; in a script, setting it to `TRUE` builds without asking, which is the
way to make a pipeline set itself up on a new machine.

To see what you have:

```{r}
TNRS_local_status()
```

```
Not built: wcvp. Add with TNRS_local_build("wcvp").
  source                          full_name built version                     doi downloaded size_mb download_mb
1   wcvp World Checklist of Vascular Plants FALSE    <NA>                    <NA>       <NA>     0.0          85
2    wfo                 World Flora Online  TRUE 2025-12 10.5281/zenodo.18007552 2026-08-30   119.9         116
```

Every source the package can use is listed, so this answers both questions at
once: what you resolved against, and what else you could build. `built` is FALSE
for a source you have not downloaded, and `version`, `doi` and `downloaded` are
then NA, since there is nothing installed to describe. `size_mb` is what the
source occupies on disk right now, so it is larger for a source built with
`keep_archive = TRUE`; `download_mb` is what fetching it would cost.

Keep the rows where `built` is TRUE. They are what makes a local result citable:
they record exactly which version of which source you resolved against.

`TNRS_local_remove()` deletes the cache if you want the space back.

## Resolving names

```{r}
results <- TNRS_local(c("Quercuss alba", "Xantium strumarium", "Acer rubrum"))

results[, c("Name_submitted", "Name_matched", "Overall_score", "Accepted_name")]
```

```
      Name_submitted        Name_matched Overall_score       Accepted_name
1      Quercuss alba        Quercus alba     0.9792205        Quercus alba
2 Xantium strumarium Xanthium strumarium     0.9792205 Xanthium strumarium
3        Acer rubrum         Acer rubrum     1.0000000         Acer rubrum
```

The output has the same columns as `TNRS()`, so code that reads the result of
one will read the other.

The first call in a session loads the backbone, which takes about twenty seconds
per source. After that it is held in memory and every later call is pure
matching.

## How much faster is it?

The table below is from a **single comparison on one machine**, resolving real
names drawn at random from the backbone. Treat it as an indication of the shape
of the difference, not as a benchmark: it will vary with your hardware, your
network, and how busy the web service happens to be when you run it. In
repeated runs here the web service's timings varied by a factor of two from one
call to the next.

| names | `TNRS()` (web service) | `TNRS_local()`, one source | `TNRS_local()`, both |
|---|---|---|---|
| 50 | 23 s | 10 s | 4 s |
| 200 | 71 s | 4 s | 5 s |
| 1000 | 376 s | 11 s | 16 s |

The web service ran at roughly three names a second whatever the batch size,
since the work happens on the server and every batch is a round trip. Locally
the rate climbs with the size of the job, to something like 60-90 names a second
for a thousand names, because the fixed costs are paid once.

**For small jobs the web service can still win.** Loading the backbone costs
about twenty seconds, which a short query does not earn back:

* 50 names, from a cold session: about 31 s locally against 23 s for the web
  service.
* 1000 names, from a cold session: about 33 s locally against 376 s.

Break-even was somewhere near seventy names in a fresh session, and beyond that
the gap widens quickly, because the local cost is nearly flat while the web
service's grows with every name. Within a session the backbone stays loaded, so
the twenty seconds is paid once however many times you call `TNRS_local()`.

If you are resolving a handful of names once, `TNRS()` is simpler and needs no
setup. If you are resolving thousands, or the same names repeatedly, or you have
no reliable connection, the local version is the better tool.

## How local results differ from the web service

They usually agree, but not always, and the differences are worth understanding
before you rely on either.

### The local sources are usually newer

The web service updates its copy of the sources periodically. `TNRS_local_build()`
downloads directly from the publishers, so your copy is typically more current.
At the time of writing the service is running World Flora Online v.2023.06 while
the current release is 2025-12.

That is usually an improvement, but it means a name can legitimately resolve
differently in the two. If you compare them, expect real taxonomic updates among
the differences, not just bugs. Report the versions from `TNRS_local_status()`
so the reader knows which vintage you used.

### One source by default, rather than two

`TNRS()` consults `wcvp` and `wfo` together. When the two disagree about whether
a name is accepted, the answer you get is assembled from both, and nothing in
the output tells you that happened.

`TNRS_local()` uses `wfo` alone by default, so the answer comes from one
authority. To reproduce the web service's behaviour, ask for both:

```{r}
both <- TNRS_local(
  c("Bartlettina macrocephala", "Acer rubrum"),
  sources = c("wcvp", "wfo")
)

both[, c("Name_submitted", "Taxonomic_status", "Accepted_name", "Source",
         "Source_conflict")]
```

```
            Name_submitted Taxonomic_status            Accepted_name Source Source_conflict
1 Bartlettina macrocephala         Accepted Bartlettina macrocephala    wfo            TRUE
2              Acer rubrum         Accepted              Acer rubrum   wcvp           FALSE
```

`Source_conflict` marks names where the sources led to different accepted names.
Those are the ones worth checking by hand. To see what each source actually
said:

```{r}
TNRS_local("Bartlettina macrocephala",
           sources = c("wcvp", "wfo"), matches = "all")
```

```
              Name_matched Taxonomic_status            Accepted_name Source
1 Bartlettina macrocephala         Accepted Bartlettina macrocephala    wfo
2 Bartlettina macrocephala          Synonym  Bartlettina ehrenbergii   wcvp
```

Neither answer is wrong. Kew and World Flora Online genuinely disagree about
this name, and a service that consults both has to pick one.

### `accuracy` filters differently

In `TNRS_local()`, `accuracy` discards matches whose overall score is below the
threshold, which is what the documentation of the web service describes.

The web service itself is more permissive: it discards a match only when
*every* component score is below the threshold. Because a genus can score highly
while the epithet fails entirely, a match with a low overall score can survive a
high threshold there.

## Names that are not plants

Neither the web service nor `TNRS_local()` knows what a plant is. Both match
against plant checklists, so a non-plant name is either unmatched or matched to
whatever plant name happens to be closest:

```{r}
TNRS_local("Dendroica palmarum")[, c("Name_matched", "Overall_score",
                                     "Accepted_name", "WarningsEng")]
```

*Dendroica palmarum* is a warbler. It can match a plant genus at a low score.

Two columns tell you when this has happened:

* `Overall_score` will be low.
* `WarningsEng` will contain `[Partial]`, meaning the match did not reach the
  rank the name was parsed to. A binomial that only matched at genus rank, with
  its epithet unmatched, is the signature of this problem.

If your data may contain non-plants, filter on those rather than taking
`Accepted_name` at face value.

## Using your own checklist

`TNRS_local_add_source()` registers a checklist you supply as a source, on the
same footing as the ones this package downloads. Use it for an authority the
TNRS does not distribute, for a group outside the flowering plants, or for an
in-house list.

The web service offers a Cactaceae checklist from caryophyllales.org as source
`cact`. That one is not published for download, so this package cannot fetch it
for you, but if you obtain a copy you can register it:

```{r}
TNRS_local_add_source(
  "cactaceae.csv",
  source = "cact",
  version = "2021",
  full_name = "Cactaceae at Caryophyllales.org",
  doi = "10.3372/wi.51.51208"
)
```

Checklists are not always published as a single Darwin Core table. Those at
Caryophyllales.org come as a workbook of two sheets, accepted taxa and synonyms,
exported from the EDIT platform. `TNRS_cdm_to_dwc()` converts that shape:

```{r}
accepted <- readxl::read_excel("CactaceaeFullList.xlsx", sheet = "Accepted")
synonym <- readxl::read_excel("CactaceaeFullList.xlsx", sheet = "Synonym")

cact <- TNRS_cdm_to_dwc(accepted, synonym, family = "Cactaceae")
TNRS_local_add_source(cact, source = "cact", version = "2023-11-02")
```

It stacks the two sheets, drops the placeholder nodes that organise the tree
rather than name a taxon, and reads each name's status off the tree, so that
names the authors could not place stay unchecked rather than becoming accepted
taxa.

Columns are otherwise looked for under their Darwin Core names, which is how
checklists are normally published. A list that names them differently needs a mapping, and
only the name itself is required:

```{r}
TNRS_local_add_source(
  my_list,
  source = "inhouse", version = "2026-09",
  columns = c(scientific_name = "taxon", authorship = "author")
)
```

Anything the checklist does not carry is worked out from the name by the same
parser the resolver uses, so a single column of names is enough to start. A
checklist with no status column is treated as a list of accepted names.

Once registered, it behaves like any other source. It can be used on its own:

```{r}
TNRS_local(c("Mamillaria elongata"), sources = "cact")
```

or alongside a global source, which is the interesting case, since a specialist
checklist is worth having precisely where it disagrees with a global one:

```{r}
results <- TNRS_local(names, sources = c("cact", "wfo"))
results[results$Source_conflict, ]
```

Two things to know. `version` is required, because a result resolved against an
unversioned local file cannot be reported reproducibly; it is recorded and shown
by `TNRS_local_status()` like any other source's. And `TNRS_local_remove()`
deletes sources you supplied along with everything else, and this package cannot
fetch them again, so keep the original file.

## Animals, and other codes of nomenclature

The TNRS is a plant service, but the local engine is not limited to plants. Two
animal-bearing sources are available:

```{r}
TNRS_local_build("mdd")   # Mammal Diversity Database, 68 MB
TNRS_local_build("col")   # Catalogue of Life, all life, 488 MB
```

Neither is built by default, and `col` in particular is by far the largest source
offered: 5.4 million names against WFO's 1.6 million. Build it only if you need
its breadth.

Sources record which nomenclatural code they follow, and `TNRS_local()` takes
the code from the sources you ask for. That matters mainly for family prefixes,
since botanical families end in `-aceae` and zoological ones in `-idae`:

```{r}
TNRS_local("Felidae Panthera leo", sources = "mdd")
```

resolves, where the same name against a plant source would not, because
`Felidae` would be read as the genus. The `nomenclature` argument overrides the
setting if you need to.

Asking for a plant and an animal source together resolves under both codes:

```{r}
TNRS_local(names, sources = c("mdd", "wfo"))
```

This is worth doing when your data may contain either. The two codes are
independent, so the same genus name can be validly used for a plant and an
animal, *Oenanthe* being both a bird and a plant genus. Resolving under one code
alone hides that; resolving under both, with `Source` telling you which backbone
each match came from, makes it visible.

If your data are entirely non-plant, a single-kingdom source is the safer
choice. Resolving animal names against a plant backbone produces confident
nonsense, which is the subject of the previous section.

## Known limitations

* **Badly misspelled names can be missed.** The matcher narrows the search
  before comparing, using the first and last few characters of the genus. A
  genus misspelled at *both* ends can therefore be missed however close it
  otherwise is. This is a property of the published TNRS algorithm, which the
  local version reproduces. In practice it affects under 1% of single-character
  errors, since real misspellings usually preserve the opening.
* **Second infraspecific epithets are matched exactly, but not fuzzily.** A name
  such as `Cirsium japonicum var. vestitum f. arakii` resolves in full when the
  source carries it. A *misspelling* of the second epithet does not: the name
  falls back to its parent taxon, which is correct as far as it goes - the forma
  does sit within the variety - but loses the sub-taxon granularity, and is
  reported as `[Partial]` in `WarningsEng`. Such names are rare: none at all in
  WCVP v15, and 44 of 1.66 million in WFO 2025-12, all of them accepted taxa.

  The web service does not resolve to this level at all. It has the
  `Infraspecific_rank_2` and `Infraspecific_epithet_2_matched` columns but never
  populates them, and for two of the three such names tested it returned no
  match at all.
* **`Name_matched_lsid` and `Accepted_name_lsid` are always empty**, because the
  sources do not supply them.

## Citing your results

Cite the TNRS itself, the taxonomic sources, and the versions you used:

```{r}
# The TNRS and its sources
TNRS_metadata()

# The versions your local results came from
TNRS_local_status()
```
