---
title: "Resolving names offline"
author: "Brian Maitner"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Resolving names offline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE)
```

## Why resolve names offline

`TNRS()` sends your names to the Taxonomic Name Resolution Service and returns
the server's answer. That is the right choice most of the time, but it needs an
internet connection, it needs the service to be up, and very large jobs are sent
in batches of 5000 with a round trip for each.

`TNRS_local()` runs the same matching algorithm on your own machine, against a
copy of the taxonomic sources that you download once. It works with no network,
it does not depend on the service being available, and there are no round trips.

The two are separate functions rather than one function with a switch. That is
deliberate: they do not always give the same answer, and it should always be
obvious from your code which one produced a result.

## One-off setup

```{r}
library(TNRS)

TNRS_local_build()
```

This downloads the taxonomic sources from their publishers and prepares them for
matching. It is roughly 200 MB and takes a few minutes. You only need to do it
again when you want newer versions of the sources.

To see what you have:

```{r}
TNRS_local_status()
```

```
  source                          full_name version                     doi downloaded size_mb
1   wcvp World Checklist of Vascular Plants     v15                    <NA> 2026-08-28    85.4
2    wfo                 World Flora Online 2025-12 10.5281/zenodo.18007552 2026-08-28   115.5
```

Keep that output. It is what makes a local result citable: it records exactly
which version of which source you resolved against.

`TNRS_local_remove()` deletes the cache if you want the space back.

## Resolving names

```{r}
results <- TNRS_local(c("Quercuss alba", "Xantium strumarium", "Acer rubrum"))

results[, c("Name_submitted", "Name_matched", "Overall_score", "Accepted_name")]
```

```
      Name_submitted        Name_matched Overall_score       Accepted_name
1      Quercuss alba        Quercus alba     0.9792205        Quercus alba
2 Xantium strumarium Xanthium strumarium     0.9792205 Xanthium strumarium
3        Acer rubrum         Acer rubrum     1.0000000         Acer rubrum
```

The output has the same columns as `TNRS()`, so code that reads the result of
one will read the other.

The first call in a session loads the backbone, which takes about twenty seconds
per source. After that it is held in memory, and matching runs at tens of
thousands of names per minute.

## How local results differ from the web service

They usually agree, but not always, and the differences are worth understanding
before you rely on either.

### The local sources are usually newer

The web service updates its copy of the sources periodically. `TNRS_local_build()`
downloads directly from the publishers, so your copy is typically more current.
At the time of writing the service is running World Flora Online v.2023.06 while
the current release is 2025-12.

That is usually an improvement, but it means a name can legitimately resolve
differently in the two. If you compare them, expect real taxonomic updates among
the differences, not just bugs. Report the versions from `TNRS_local_status()`
so the reader knows which vintage you used.

### One source by default, rather than two

`TNRS()` consults `wcvp` and `wfo` together. When the two disagree about whether
a name is accepted, the answer you get is assembled from both, and nothing in
the output tells you that happened.

`TNRS_local()` uses `wfo` alone by default, so the answer comes from one
authority. To reproduce the web service's behaviour, ask for both:

```{r}
both <- TNRS_local(
  c("Bartlettina macrocephala", "Acer rubrum"),
  sources = c("wcvp", "wfo")
)

both[, c("Name_submitted", "Taxonomic_status", "Accepted_name", "Source",
         "Source_conflict")]
```

```
            Name_submitted Taxonomic_status            Accepted_name Source Source_conflict
1 Bartlettina macrocephala         Accepted Bartlettina macrocephala    wfo            TRUE
2              Acer rubrum         Accepted              Acer rubrum   wcvp           FALSE
```

`Source_conflict` marks names where the sources led to different accepted names.
Those are the ones worth checking by hand. To see what each source actually
said:

```{r}
TNRS_local("Bartlettina macrocephala",
           sources = c("wcvp", "wfo"), matches = "all")
```

```
              Name_matched Taxonomic_status            Accepted_name Source
1 Bartlettina macrocephala         Accepted Bartlettina macrocephala    wfo
2 Bartlettina macrocephala          Synonym  Bartlettina ehrenbergii   wcvp
```

Neither answer is wrong. Kew and World Flora Online genuinely disagree about
this name, and a service that consults both has to pick one.

### `accuracy` filters differently

In `TNRS_local()`, `accuracy` discards matches whose overall score is below the
threshold, which is what the documentation of the web service describes.

The web service itself is more permissive: it discards a match only when
*every* component score is below the threshold. Because a genus can score highly
while the epithet fails entirely, a match with a low overall score can survive a
high threshold there.

## Names that are not plants

Neither the web service nor `TNRS_local()` knows what a plant is. Both match
against plant checklists, so a non-plant name is either unmatched or matched to
whatever plant name happens to be closest:

```{r}
TNRS_local("Dendroica palmarum")[, c("Name_matched", "Overall_score",
                                     "Accepted_name", "WarningsEng")]
```

*Dendroica palmarum* is a warbler. It can match a plant genus at a low score.

Two columns tell you when this has happened:

* `Overall_score` will be low.
* `WarningsEng` will contain `[Partial]`, meaning the match did not reach the
  rank the name was parsed to. A binomial that only matched at genus rank, with
  its epithet unmatched, is the signature of this problem.

If your data may contain non-plants, filter on those rather than taking
`Accepted_name` at face value.

## Known limitations

* **Badly misspelled names can be missed.** The matcher narrows the search
  before comparing, using the first and last few characters of the genus. A
  genus misspelled at *both* ends can therefore be missed however close it
  otherwise is. This is a property of the published TNRS algorithm, which the
  local version reproduces. In practice it affects under 1% of single-character
  errors, since real misspellings usually preserve the opening.
* **Second infraspecific ranks are not matched.** A name carrying two
  infraspecific epithets, such as `Cirsium japonicum var. vestitum f. arakii`,
  resolves to its parent taxon (here the variety). That is a correct answer as
  far as it goes - the forma does sit within the variety - but you lose the
  sub-taxon granularity, so it is reported as `[Partial]` in `WarningsEng`.
  Such names are rare: none in WCVP v15, and 44 of 1.66 million in WFO 2025-12.
* **`Name_matched_lsid` and `Accepted_name_lsid` are always empty**, because the
  sources do not supply them.

## Citing your results

Cite the TNRS itself, the taxonomic sources, and the versions you used:

```{r}
# The TNRS and its sources
TNRS_metadata()

# The versions your local results came from
TNRS_local_status()
```
